Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-16 doc/lean.tex
describe model counting in Lean

\subsection{The Proof Checker}

@@

[TODO(JA): I'll leave stop here now because I am running out of steam today,
and also because Wojciech is still working on it.
But here is where we can sketch what the checker does.
For each line, it does some checks, it adds clauses to a clause database,
and it add a definition to the POG.
There is a list of invariants having to do with the original formula,
the clause database, and the POG, and we prove that if the checks are satisfied,
then at each step the invariants hold.
We should list the invariants and say something about how the checks are
performed.]

@@

The number of models that a propositional formula has depends on what we take the
variables to be. For example, there is only one assignment to the variables
$\{ x, y \}$ that makes $x \wedge y$ true, but there are two assignments to the
variables $\{ x, y, z \}$ that make $x \wedge y$ true.
It is not hard to show that for partitioned formulas \lstinline{PropForm ν} whose
variables are among a finite set \lstinline{s} of variables of cardinality \lstinline{numVars},
we can count the number of models of the formula on the variables in \lstinline{s}
recursively as follows:
\begin{lstlisting}
def countModels (nVars : Nat) : PropForm ν → Nat
  | tr         => 2^nVars
  | fls        => 0
  | var _      => 2^(nVars - 1)
  | neg φ      => 2^nVars - φ.countModels nVars
  | disj φ ψ   => φ.countModels nVars + ψ.countModels nVars
  | conj φ ψ   => φ.countModels nVars * ψ.countModels nVars / 2^nVars
  | impl _ _   => 0
  | biImpl _ _ => 0
\end{lstlisting}
Note that we can carry out this calculation for any formula at all, though the
calculation is meaningless if the formula is not partitioned.
Also recall that if the formula is partitioned, it does not contain implication or bi-implication.

We have implemented an efficient counting function for POGs and proved that
it computes that same quantity on the corresponding formulas:
\begin{lstlisting}
def count (pog : Pog) (nVars : Nat) (x : Var) : Nat := ...

theorem count_eq_countModels (pog : Pog) (nVars : Nat) (x : Var) :
  pog.count nVars x = (pog.toPropForm (.mkPos x)).countModels nVars
\end{lstlisting}
Here \lstinline{.mkPos x} denotes the positive literal for the variable \lstinline{x}. We have also proved that the calculation above really does
return the number of models:
\begin{lstlisting}
theorem countModels_eq_card_models {φ : PropForm ν} {s : Finset ν}
    (hvars : φ.vars ⊆ s) (hpar : φ.partitioned) :
  φ.countModels s.card = card (φ.models s)
\end{lstlisting}
In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
we have that the number of models on its variables is \lstinline{φ.countModels φ.vars.card}.

Given a weighting function \lstinline{weight : ν → Rat} that assigns a rational
number in $[0, 1]$ to each variable, we can calculate the weighted model count on
partitioned formulas as follows:
\begin{lstlisting}
def countWeight (weight : ν → Rat) : PropForm ν → Rat
  | tr         => 1
  | fls        => 0
  | var x      => weight x
  | neg φ      => 1 - φ.countWeight weight
  | disj φ ψ   => φ.countWeight weight + ψ.countWeight weight
  | conj φ ψ   => φ.countWeight weight * ψ.countWeight weight
  | impl _ _   => 0
  | biImpl _ _ => 0
\end{lstlisting}
We have implemented the corresponding count function on POGs and proved that
it calculates this quantity on the associated formulas,
though we have not yet proved the combinatorial lemma
that relates this calculation to the sum of the weights of the corresponding models.

@@

In this subsection, we clarify what has been verified and what has been trusted.
Recall that our first step is to parse a CRAT file to read in the initial CNF formula
and the CRAT proof. We do not verify this step. There is no strong need to verify the
parsing of the proof, because if the checker claims success for \emph{any} proof, it does not really matter
whether the proof was the one the user intended.
On the other hand, it would be nice to have reassurance that the CNF formula that serves
as input to our verified checker matches the one described in the file.
An easy and fairly standard way of addressing this is to print out the CNF formula
right after parsing it and then do a diff with the original input.
This involves trusting only the correctness of the print procedure and diff.

Lean's code extraction replaces calculations with natural numbers and integers with
efficient but unverified arbitrary precision versions.
Lean also uses an efficient implementation of arrays; within the
formal system, these are defined in terms of lists, but code extraction replaces them
with dynamic arrays and uses reference counting to allow destructive updates when it is safe
to do so [cite counting immutable beans paper].
Code extraction also provides efficient implementations of tail recursive functions and loops
[cite ``do unchained'' paper].
Finally, we have implemented hashmaps in terms of arrays, but we have not yet proved
the correctness of all the primitive operations on these hashmaps.
Thus our proofs depend on the unverified assumption that those operations have the
expected properties.

@@

[TODO: Say something to establish that the procedure run in reasonable time and produce useful results; or maybe this does in the next section?]

>>>>

\subsection{The Proof Checker}
\label{subsection:proof:checker}

@@

@@

We have formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem,
Definition~\ref{def:ring_evaluation}, as follows:
\begin{lstlisting}
def weightCount {R : Type} [CommRing R]
    (weight : ν → R) (φ : PropForm ν) (s : Finset ν) : R :=
  ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
\end{lstlisting}
Here \lstinline{R} is assumed to be a commutative ring. The curly brackets around \lstinline{R} indicate that
this argument is meant ot be left implicit, because it can inferred from the return type of
the weight function. Similarly, the square brackets indicate that the commutative ring structure
is a type class argument, generally inferred from the context.

More precisely, the quantity \lstinline{weightCount weight φ s} adds up the weights of the models
of \lstinline{φ} over the set of variables \lstinline{s}. We prove that as long as \lstinline{s}
contains the variables of \lstinline{φ}, adding extra variables does not change the sum, that is, \lstinline{vars φ ⊆ s} implies
\begin{lstlisting}
  weightCount weight φ s = weightCount weight φ (vars φ)
\end{lstlisting}
The counting scheme of Proposition~\ref{prop:ring:eval} for partitioned formulas is expressed as follows:
\begin{lstlisting}
 def countWeight (weight : ν → Rat) : PropForm ν → Rat
  | tr       => 1
  | fls      => 0
  | var x    => weight x
  | neg φ    => 1 - countWeight weight φ
  | disj φ ψ => countWeight weight φ + countWeight weight ψ
  | conj φ ψ => countWeight weight φ * countWeight weight ψ
\end{lstlisting}
We arbitrarily assign 0 to implications and bi-implications;
recall that, by definition, these do not appear in partitioned formulas.
Proposition~\ref{prop:ring:eval} is then formalized as follows:
\begin{lstlisting}
theorem countWeight_eq_weightCount (weight : ν → R) {φ : PropForm ν}
    (hdec : partitioned φ) :
  countWeight weight φ = weightCount weight φ (vars φ)
\end{lstlisting}

The functions just described are not meant to be computed, but,
rather, serve as our mathematical reference.
We have implemented an efficient function to calculate the weighted model count
for an arbitrary node of a Pog, and we have proved that it computes the function above on the
associated formula:
\begin{lstlisting}
theorem countWeight_eq_countWeight (pog : Pog) (weight : Var → Rat)
    (x : Var) :
  pog.countWeight weight x = (pog.toPropForm (.mkPos x)).countWeight weight
\end{lstlisting}
Here \lstinline{.mkPos x} denotes the positive literal for the variable \lstinline{x}.
Applying this to the output of our verified CRAT proof checker,
we obtain a proof that
our toolchain computes the correct weighted model count of the input CNF.

Lean 4 does not yet have an efficient implementation of rational arithmetic, and
we have not yet implemented a verified version of $\drational$,
defined in Section~\ref{sect:experimental}.
We have therefore implemented a separate procedure that carries out an integer
calculation of the number of models.
The number of models that a propositional formula has depends on what we take the
variables to be. For example, there is only one assignment to the variables
$\{ x, y \}$ that makes $x \wedge y$ true, but there are two assignments to the
variables $\{ x, y, z \}$ that make $x \wedge y$ true.
It is not hard to show that for partitioned formulas \lstinline{PropForm ν} whose
variables are among a finite set \lstinline{s} of variables of cardinality \lstinline{numVars},
we can count the number of models of the formula on the variables in \lstinline{s}
recursively as follows:
\begin{lstlisting}
def countModels (nVars : Nat) : PropForm ν → Nat
  | tr       => 2^nVars
  | fls      => 0
  | var _    => 2^(nVars - 1)
  | neg φ    => 2^nVars - countModels nVars φ
  | disj φ ψ => countModels nVars φ + countModels nVars ψ
  | conj φ ψ => countModels nVars φ * countModels nVars ψ / 2^nVars
\end{lstlisting}
We have also implemented an efficient counting function for POGs and proved that
it computes that same quantity on the corresponding formulas.
Finally,
We have proved that the calculation above really does
return the number of models:
\begin{lstlisting}
theorem countModels_eq_card_models {φ : PropForm ν} {s : Finset ν}
    (hvars : vars φ ⊆ s) (hpar :  partitioned φ) :
  countModels (card s) φ = card (models φ s)
\end{lstlisting}
In particular, taking \lstinline{s} to be exactly the variables of \lstinline{φ},
we have that the number of models on its variables is \lstinline{countModels φ (card (vars φ))}.

\subsection{Implementation of the Proof Checker}

We have specified the behavior of the proof checker in Section~\ref{subsection:proof:checker};
in this section we sketch the implementation.
Establishing its correctness was, by far, the most challenging part of the formalization
process.
For clarity, we describe the algorithm with ordinary mathematical notation rather than
Lean snippets.

We begin by reading the input CNF formula $\varphi$ and adding its clauses to a clause
database $C$. We initialize that POG graph $G$ with the variables of $\varphi$,
and then iteratively process each line of the CRAT proof as follows:
\begin{itemize}
  \item (Summarize what each step does.)
\end{itemize}
We establish that all the following invariants hold at each step:
\begin{itemize}
  \item (Wojciech has a long list of invariants)
\end{itemize}
At the end \ldots [Explain how the invariants at the end imply the desired conclusion.]

@@

In this subsection, we clarify what has been verified and what has been trusted.
Recall that our first step is to parse a CRAT file to read in the initial CNF formula
and the CRAT proof. We do not verify this step.
If one is only interested in obtaining a model count or a verified model count,
there is no strong need to verify the parsing of the proof,
because if the checker succeeds in constructing an equivalent POG,
it does not really matter whether the proof was the one the user intended.
But it would be nice to have reassurance that the CNF formula that serves
as input to our verified checker matches the one described in the file.
An easy and fairly standard way of addressing this is to print out the CNF formula
right after parsing it and then do a diff with the original input.
This involves trusting only the correctness of the print procedure and diff.
Similarly, if one wants to establish the correctness of the POG contained in the CRAT file,
one can print out the POG that is constructed by the checker and compare.

Lean's code extraction replaces calculations with natural numbers and integers with
efficient but unverified arbitrary precision versions.
Lean also uses an efficient implementation of arrays; within the
formal system, these are defined in terms of lists, but code extraction replaces them
with dynamic arrays and uses reference counting to allow destructive updates when it is safe
to do so [cite counting immutable beans paper].
Code extraction also provides efficient implementations of tail recursive functions and loops
[cite ``do unchained'' paper].
Finally, Lean's standard library implements hashmaps in terms of array.
Many of the basic properties of hashmaps have been formally verified, but not all.
In particular, we make use of a fold operation whose verification is not yet complete.
Thus our proofs depend on the assumption that the fold operation has the
expected properties.

@@

[TODO: Say something to establish that the procedure run in reasonable time and produce
useful results; or maybe this goes in the next section?]
